Helixer: Cross-species Gene Annotation Of Large Eukaryotic Genomes Using Deep Learning

### 2020-07-20 paper/main.tex
co-author suggestions and just little adjustments from re-reading

\title[short Title]{Helixer: Cross-species Gene Annotation Of Large Eukaryotic Genomes Using Deep Learning}
\author[Sample \textit{et~al}.]{Felix Stiehler\,$^{\text{\sfb 1}}$, Marvin Steinborn\,$^{\text{\sfb 1}}$, Stephan Scholz, Daniela Dey, Andreas P.M. Weber\,$^{\text{\sfb 1}}$, Alisandra K. Denton\,$^{\text{\sfb 1,}*}$}
\address{$^{\text{\sf 1}}$Institue for Plant Biochemistry, Heinrich-Heine-University, Dusseldorf, D-40225, Germany}

@@

\section{Introduction}
The annotation of genes in DNA is fundamental for many downstream tasks, which 
depend on high quality annotations. Gene annotation can be performed at different levels 
of precision, from simple coding -- non-coding classification to detailed structural 
labeling. Because of the sheer size of genomes alone, manual gene annotation is 
generally intractable. Instead, researchers can use pipelines such as 
Maker \citep{cantarel2008maker}, PASA \citep{haas2003improving} or those 
offered by genomic database providers like NCBI 
\citep{thibaud2013eukaryotic} and Ensembl \citep{aken2016ensembl}. 
These pipelines integrate experimental data (from e.g. RNA-seq 
or proteogenomics) with homologous sequences in the database and ab-initio gene 
predictions. The latter is an attractive approach, because it is cheap and fast. 
State-of-the-art performance is achieved by higher order hidden markov models 
(HMMs), such as Genscan \citep{burge1997prediction}, AUGUSTUS \citep{stanke2003gene} 
or SNAP \citep{johnson2008snap}. Their accuracy, however, leaves room for 
improvement. By encoding possible states and transitions in a probalistic model, 
designers of HMMs assume structure in the sequence that may limit its predictive 
power. In practice HMMs have trouble generalizing across species and any actual learning 
related to sequence motifs is limited to very short sequences that indicate state transitions.

In the last decade, deep neural networks (DNN) have been applied with great success 
in many areas of statistical modelling, including biology 
\citep{ching2018opportunities}. For sequence data, such as DNA, speech or text, a 
special kind of recurrent neural network (RNN) called long-short term memory (LSTM) 
\citep{hochreiter1997long} is an established building block for many different 
architectures. LSTM units can also be used to process sequential input starting 
from both ends, forming a bi-directional LSTM (BLSTM). It has also been shown that 
HMMs can be successfully combined with DNNs \citep{liu2016novo, liu2016pedla}.

For the purpose of gene annotation, RNNs have already shown promising results. 
\citep{choudharypredicting} carries out preliminary explorations on the potential 
of BLSTMs for cross species gene prediction and trains his model on human genes to 
test it later on two more species. DeepAnnotator \citep{amin2018deepannotator} uses 
bidirectional LSTMs (BLSTM) for gene finding in prokaryotes. Gene prediction in 
prokaryotes is considered more amenable than in eukaryotes, as genes in prokaryotes 
are proportionately more frequent in the genome, feature simpler control structures 
and do not use splicing \citep{wang2004brief}. DanQ \citep{quang2016danq} proposes 
the use of a BLSTM after a convolutional neural net (CNN) to find detailed motifs in 
the human genome. DeePromoter \citep{oubounyt2019deepromoter} trains a similar 
architecture for the recognition of promoter regions. Recently, several groups 
\citep{jaganathan2019predicting, wang2019splicefinder} successfully used CNNs to 
find splicing sites.

In this work we present Helixer, a novel prototype software for training and 
utilizing a general purpose DNN for the ab-initio cross-species gene annotation of 
large eukaryotic genomes using only DNA sequence as input. Our model is trained to 
differentiate between 4 regions: Intergenic, Untranslated (UTR), Coding (CDS) and 
Intron. We demonstrate the effectiveness of this approach by training a single model 
for the annotation of a large set of genomes from the kingdoms metazoa and 
viridiplantae, respectively, which we will call {\it animals} and {\it plants} from 
now on. We worked with the full data of 192 animal genomes and 60 plant genomes. 
Both the ability to generalize across genomes as well as the scope of the evaluation 
greatly surpasses all other work in this field. The source code and all input data 
is publicly available.

@@

For the selection of training genomes we had to balance multiple conflicting 
trade-offs. On the one hand, we want to train with as much data as possible, but 
on the other
not all data has sufficient quality to enable a powerful generalization. It is 
also very desirable to have diverse training genomes with a broad phylogenetic spread, 
a variety of genome sizes and average gene lengths as the model is increasingly unlikely to generalize 
well beyond the borders given by the training data. However, it also may be difficult 
for the model learn and generalize if the genomic patterns inside the data are 
too different from each other. Practically, it was important for us to limit the size of
our training genomes set to be small enough that enough that we could get experimental 
results within a couple of days and thereby be able to test many different data and 
model configurations.

We employed an iterative approach to effectively select a proper set of training 
genomes. We first started with 3-4 genomes that were expected to be of the 
highest quality. We then evaluated the performance of the resulting model on all 
genomes individually (including the training genomes) and looked for candidates to 
add to the training set or to remove from it. This process was repeated multiple 
times concurrently with the model search until we saw no more room for substantial 
improvement given our computational constraints.

We call the genomes that were ultimately not used during training from the original 
split the {\it validation genomes}. While we did split off a group of test genomes 
at the beginning of our work, we ended up only reporting on the generalization 
performance on the combined set of validation and test genomes as the validation 
genomes are by far the largest set. We also did not find a noticeable difference in 
performance between those two groups.

@@

\subsection{Data Pre-processing}
We first pre-process and store the raw genomic information by using GeenuFF 
(https://github.com/weberlab-hhu/GeenuFF). 
GeenuFF is a tool for checking and exploring genomic data, that 
stores all information inside a SQL database. Training and evaluation-ready data is 
generated by querying this database and then transforming the returned data into a 
numerical format suitable for machine learning. The encoding of the genomic sequence 
is done in line with the IUPAC nucleic acid notation and the functional annotation 
used as labels during training is transformed to a one hot encoding with the four 
classes {\it Intergenic}, {\it UTR}, {\it CDS} and {\it Intron}. See also Table 
\ref{tab:arrays} for a more detailed description of the generated data types.

@@

During data generation, we query for the transcript with the longest protein of each gene and 
disregard FASTA sequences that have no functional annotation as it is ambiguous
whether such sequences contain no genes or were simply not annotated in the reference. GeenuFF checks the 
genomic annotations for potential errors during import and is able to mark those 
areas. We use this information to effectively blank out those bases during training 
by using the sample weights described in Table \ref{tab:arrays}. The vast majority 
of masked bases lie in the intergenic region as the most prevalent error is a 
missing UTR and GeenuFF marks a potentially large intergenic region for it. Table 
\ref{tab:statistics_all} shows statistics about the masking.

For the training itself, we divide each continuous genomic sequence into 20,000 bp 
long subsequences, for which one-hot vectors of the base pairs and annotations 
are generated and respectively used as input and label for the neural network 
together with the sample weights. We append zero padding if the subsequences  
are shorter than 20,000 bp. 
If a subsequence is fully marked as erroneous, we exclude it from all analyses.

@@

\subsection{Metrics}
\label{sec:metrics}
We mainly use two metrics to judge model performance. The {\it Genic F1} is our 
primary metric that provides the most comprehensive picture of annotation quality 
in one number. The {\it Subgenic F1} is similar to the Genic F1, except that it does 
not take UTR predictions into account. This is done for comparability with AUGUSTUS 
as further explained in Section \ref{sec:augustus}. Both metrics work by summing up 
True Positives (TP), False Positives (FP) and False Negatives (FN) of each 
considered class before calculating precision and recall from these values and 
combining the two into a F1 score. Neither metric takes the values of the 
intergenic class into account, as this class is very abundant and appears to be by 
far the easiest to predict. (See Table \ref{tab:statistics_all} for the class 
distribution and Tables S\ref{suptab:generalization_animals} 
and S\ref{suptab:generalization_plants} for a more in-depth report our model performance 
including the intergenic class). In the case of the Genic F1, this means that 
effectively only the TP of the intergenic class are disregarded. The metrics 
essentially provide a weighted mean of the performances in the considered classes,
with weights proportional to class frequency.

@@

\subsection{Model Architecture and Training}
\label{sec:model}
We use a 4-layer deep stacked BLSTM network with 256 units per layer and layer 
normalization \citep{ba2016layer} between each BLSTM layer to produce the 
predictions in the form of a base pair wise classification. The model consists of 
circa 5.4 million parameters and was implemented with the deep learning library 
Keras (https://keras.io) on top of TensorFlow \citep{abadi2016tensorflow}. We 
tested multiple different architectures before arriving at this model configuration, 
including convolutional neural networks (CNN) and hybrid architectures. We also use 
class weights as the class frequencies are both unbalanced and 
vary greatly between animals and plants. 
For example, the trade-off between the number of intergenic and intronic bases is 
quite different in both groups (see Table \ref{tab:statistics_all}).

A class wise evaluation including the calculation of the Genic 
F1 score is performed on the validation set after every epoch. The best model
from any training run was selected as the model with the highest Genic F1 after either
maximum epochs were reached or Genic F1 stopped improving and training was interrupted.

@@

We also compare our final models to a dilated CNN (dCNN) baseline as this kind of architecture is 
widely used when working with DNA data as input. Details of the neural architecture 
search are given in Section \ref{ssec:dcnn} of the Supplements.

\subsection{Inference Techniques}
We also used multiple techniques to improve the prediction quality after the 
training was done. One very effective way for genomes with larger genes was to input 
longer sequences during inference time than during training. This was done for all 
animal genomes except the invertebrates. The input sequences were up to 10 times 
longer, depending on the phylogenetic group and assembly quality. This also demonstrates the 
ability of our model to generalize as it is able to make successful prediction on 
far longer sequences than it has ever seen. The 
concrete lengths were chosen to keep the typical average gene length roughly 
proportional to the length of the sequence input. For more implementation details 
on this see Section \ref{ssec:longer} of the Supplements or the source code.

The final predictions of a single model are constructed by overlapping predictions, 
that were made from a sliding window and then cropped to a core sequence. This was done 
to strongly reduce a typical drop in performance of the models towards the beginning 
and end of each sequence (see Figure \ref{fig:main_results}c). It also improves the 
average model performance in general by providing the model with multiple different 
starting points. The different overlapping sequences were combined by averaging 
the individual softmax values of each base. The figures in Section \ref{ssec:overlapping} of the 
Supplements show the effect of overlapping for each genome, ordered by N75. We found 
that overlapping tends to work best if the genomes are not very fragmented and we 
used it for both animals and plants.

A model ensemble with 8 components was used to generate the final predictions for 
each species. For this, we performed 4 separate animal and plant training runs and 
selected two checkpoints each, specifically the checkpoint from the epoch with the best precision and the  % please check / epoch model / version something
other from the epoch with the best recall. The model checkpoint with the best Genic F1 was always best in one of 
those. This was done to increase the diversity of the model ensemble. As with the 
overlapping, the fusion of the 8 individual predictions was done by averaging the 
softmax values of each base pair prediction.

\subsection{Evaluation of AUGUSTUS} 
\label{sec:augustus}
Due to the practical time constraints for retraining and running, we decided to compare our Helixer
models only to one existing {\it de novo} gene caller, the popular gene caller, AUGUSTUS \citep{stanke2003gene}.
Typical usage requires retraining AUGUSTUS for each species, with the exception of a
few lucky cases where a model for a sufficiently close relative is already available. To scale this for 
the large plant and animal datasets, we used protein-homology to create a training set and
therefore could train and evaluate only models without UTRs.

Orthologs of highly conserved generally single-copy genes were identified in each genome using
BUSCO. The {\it viridiplantae} set was used for plants and the {\it metazoa} set for animals.
The training genbank files were generated directly by BUSCO, by utilizing the ``--long"  % inline code?
parameter. For plants, the entire retraining could be performed as above; however, for animals
we randomly selected only half the BUSCO-generated training set, which resulted in a training set 
size and runtime comparable to the plants (about 400 genes and several days per species). For animals
the training was carried out with the subsetted training file and using the following scripts provided
by AUGUSTUS. An untrained model was setup with ``new\_species.pl", and the model was fit by running
"etraining" before and after the major hyper-parameter optimization with ``optimize\_augustus.pl".
Using the trained model for each species, we ran the main prediction ("augustus") with ``--UTR=off"
and ``--gff3=on". The gffs produced by AUGUSTUS were imported into GeenuFF and exported as HDF5
files in the same manor as the reference, allowing for the direct comparison with both the reference and 
Helixer predictions. While still feasible for some 237 species, this method allowed neither training
nor prediction of UTR regions with AUGUSTUS, so the metric Subgenic F1, which disregards the class UTR,
was used for all comparisons between AUGUSTUS and Helixer (see Section \ref{sec:metrics} for more details) .

@@

For each of these six species the following search was performed on Sequence Read Archive \citep{leinonen2010sequence}
`(("<species name>"[Organism] OR <species name>[All Fields]) AND stranded[All Fields]) 
AND ("biomol rna"[Properties] AND "library layout paired"[Properties])'. In so far as 
more than 50 samples were identified and every Nth sample was selected so that under 50
samples were chosen for further processing (see Table S\ref{suptab:rnaseq_samples}).
Each sample was prepped, mapped, and quality controlled in a pipeline using Trimmomatic 
\citep{bolger2014trimmomatic}, 
Hisat2 \citep{kim2019graph}, Samtools \citep{li2009sequence}, PicardTools 
(http://broadinstitute.github.io/picard/), 
FastQC 
(www.bioinformatics.babraham.ac.uk/projects/fastqc/),
and MultiQC (\cite{ewels2016multiqc}; see
Table S\ref{suptab:bioinfo_params} for details). The above was 
automated and the results visualized with the code here (https://github.com/weberlab-hhu/RNAsleek).
The samples were filtered to those with relatively high mapping rates, high mapping to
exonic relative to non-exonic regions, low 3' bias, normal FastQC and Trimmomatic stats, 
and a stranded mapping pattern (2nd read is sense strand). If more remained, seven
of the high quality samples were selected randomly.

@@

\begin{figure*}[!tpb]
\label{fig:main_results}
\centerline{\includegraphics[width=\textwidth]{images/main_results}}
\caption{Main overall results and other investigations. The top row shows the animal
 data; the bottom row is concerned with the plants. Orange dots are for AUGUSTUS and 
the blue ones are for Helixer. {\bf a)} Swarm - and boxplots showing the Subgenic F1 
scores for all genomes that were not used during training of Helixer, which is 
equivalent to all validation and test genomes. Each dot represents the prediction 
performace for one genome. The median prediction scores are shown by the red line. 
Details of how AUGUSTUS was used are given in Section \ref{sec:augustus}. {\bf b)} 
Scatterplots showing the prediction performances measured in Subgenic F1 by genome 
length. The regression line is shown with a 95\% confidence interval. {\bf c)} The 
average Genic F1 (in red) and basepair wise accuracy (in purple) with respect to 
the position in the 20,000 basepair long input sequences. The dashed line shows the 
same without overlapping. Each value is the average performance of a 200 basepair 
long subsection by one of the models of the final ensemble. 
See Section \ref{ssec:overlapping} of the 
Supplements for the effects of overlapping in individual species.}
\end{figure*}

@@

\section{Results}
Figure \ref{fig:main_results}a shows the side-by-side comparison of the distribution 
of performances on all non-training genomes in the animal and plant case by Subgenic 
F1. In the case of Helixer, these scores represent cross-species predictions. We 
also compare the median performances of different configurations of Helixer with 
AUGUSTUS and a dilated CNN baseline in Table \ref{tab:results}. The results show a 
clear improvement over the AUGUSTUS both in median performance and spread. We also 
significantly outperform a baseline constructed of a dilated CNN architecture.

Our models, however, tend to perform less well for the very smallest and largest 
genomes or species that are phylogenetically the furthest away from our training 
genomes. This is the case for both animals and plants and is visualized by the 
Figures S\ref{supfig:phylogeny_f1_animals} and S\ref{supfig:phylogeny_f1_plants}. 
We do not, for example, consistently predict very well on the 
algae nor on the non-avian reptiles. While one of the algae were included 
in our training genomes it accounted for a tiny proportion of the total training 
data and the non-avian reptiles were not included at all.

@@

Two techniques were used during inference to improve performance and limit model 
bias. The usage of longer input lengths helped especially for genomes that tend to 
have longer genes and was enabled by our model architecture being a relatively 
simple LSTM stack without any fully connected layers on top. We also constructed the 
final predictions out of a overlapping ones, which greatly helped to reduce 
prediction bias in most genomes. To our knowledge, neither technique has been used 
before in a model developed for gene annotation.

@@

To make an independent qualitative and quantitative evaluation of
performance we compared Helixer predictions, AUGUSTUS predictions and the
reference to RNAseq coverage data. RNAseq evaluation was performed for
three each animals and plants. RNAseq coverage provides support for an 
exonic annotation (UTR or CDS), spliced coverages provides support for an 
annotation of intron, and neither coverage nor spliced coverage is expected 
for intergenic annotations. Looking at selected subsequences (Figure \ref{fig:cov_example_main}) % and supplemental figures.
we identified cases where RNAseq supported a) both the annotations of the reference 
and Helixer, b) neither, c) the reference, but not Helixer,
and d) Helixer, but not the reference.

@@

Helixer models also sometimes showed uncertainty for larger regions.
In some cases, where Helixer did not receive RNAseq support for its highest probability
annotation, it assigned a low but non-trivial probability to the RNAseq-supported 
exon/intron pattern (Figures \ref{fig:cov_example_main}b-c, S\ref{supfig:cov_example_02}c). However, in the extreme,
there are cases where the Helixer model exhibits substantial indecision or confusion 
and shifts gradually between classes with no single class receiving a high probability 
for extended stretches (Figures S\ref{supfig:cov_example_02}c-d). 
Notably, in one of the examples (Figure S\ref{supfig:cov_example_02}d)
the RNAseq shows evidence of alternative splicing; 
and in another (Figure S\ref{supfig:cov_example_02}c) Helixer's
prediction falls between that of the reference and of AUGUSTUS.

>>>>

\title[short Title]{Helixer: Cross-species Gene Annotation Of Large Eukaryotic Genomes Using Deep Learning}
\author[Sample \textit{et~al}.]{Felix Stiehler\,$^{\text{\sfb 1}}$, Marvin Steinborn\,$^{\text{\sfb 1}}$, Stephan Scholz, Daniela Dey\,$^{\text{\sfb 2}}$, Andreas P.M. Weber\,$^{\text{\sfb 1}}$, Alisandra K. Denton\,$^{\text{\sfb 1,}*}$}
\address{$^{\text{\sf 1}}$Institue for Plant Biochemistry, Heinrich-Heine-University, Dusseldorf, D-40225, Germany\\
$^{\text{\sf 2}}$Institute of Human Genetics, Medical Faculty, RWTH Aachen University, Aachen, 52062, Germany}

@@

\section{Introduction}
Annotating genes is an integral part of
genomic DNA sequence analysis, with many downstream taks
dependent on annotation quality.
Gene annotation can be performed at different levels 
of precision, from simple coding -- non-coding classification to detailed structural 
labeling. Because of the sheer size of genomes alone, manual gene annotation is 
generally intractable. Instead, researchers can use pipelines such as 
Maker \citep{cantarel2008maker}, PASA \citep{haas2003improving} or those 
offered by genomic database providers like NCBI 
\citep{thibaud2013eukaryotic} and Ensembl \citep{aken2016ensembl}. 
These pipelines integrate experimental data (from e.g. RNA-seq 
or proteogenomics) with homologous sequences in the database and ab-initio gene 
predictions. The latter is an attractive approach, because it is cheap and fast. 
State-of-the-art performance is achieved by higher order hidden markov models 
(HMMs), such as Genscan \citep{burge1997prediction}, AUGUSTUS \citep{stanke2003gene} 
or SNAP \citep{johnson2008snap}. Their accuracy, however, leaves room for 
improvement. By encoding possible states and transitions in a probalistic model, 
designers of HMMs assume structure in the sequence that may limit its predictive 
power. In practice HMMs have trouble generalizing across species and the actual learning 
of sequence motifs is limited to very short sequences that indicate state transitions.

In the last decade, deep neural networks (DNN) have been applied with great success 
in many areas of statistical modelling, including biology 
\citep{ching2018opportunities}. For sequence data, such as DNA, speech or text, a 
special kind of recurrent neural network (RNN) called long-short term memory (LSTM) 
\citep{hochreiter1997long} is an established building block for many different 
architectures. LSTM units can also be used to process sequential input starting 
from both ends, forming a bidirectional LSTM (BLSTM). It has also been shown that 
HMMs can be successfully combined with DNNs \citep{liu2016novo, liu2016pedla}.

For the purpose of gene annotation, RNNs have already shown promising results. 
\citep{choudharypredicting} carries out preliminary explorations on the potential 
of BLSTMs for cross species gene prediction and trains his model on human genes to 
test it later on two more species. DeepAnnotator \citep{amin2018deepannotator} uses 
BLSTMs for gene finding in prokaryotes. Gene prediction in 
prokaryotes is considered more amenable than in eukaryotes, as genes in prokaryotes 
are proportionately more frequent in the genome, feature simpler control structures 
and do not use splicing \citep{wang2004brief}. DanQ \citep{quang2016danq} proposes 
the use of a BLSTM after a convolutional neural net (CNN) to find detailed motifs in 
the human genome. DeePromoter \citep{oubounyt2019deepromoter} trains a similar 
architecture for the recognition of promoter regions. Recently, several groups 
\citep{jaganathan2019predicting, wang2019splicefinder} successfully used CNNs to 
find splicing sites.

In this work we present Helixer, a novel prototype software for training and 
utilizing a general purpose DNN for the ab-initio cross-species gene annotation of 
large eukaryotic genomes using only DNA sequence as input. Our model is trained to 
differentiate between 4 regions: Intergenic, Untranslated (UTR), Coding (CDS) and 
Intron. We demonstrate the effectiveness of this approach by training a single model 
for the annotation of a large set of genomes from the kingdoms metazoa and 
viridiplantae, respectively, which we will call {\it animals} and {\it plants} from 
now on. We worked with the full data of 192 animal genomes and 60 plant genomes. 
Both the ability to generalize across genomes as well as the scope of the evaluation 
represent cutting edge progress in this field. The source code and all input data 
is publicly available.

@@

For the selection of training genomes we had to balance multiple conflicting 
trade-offs. On the one hand, we want to train with as much data as possible, but 
on the other
not all data has sufficient quality to enable a powerful generalization. It is 
also very desirable to have diverse training genomes with a broad phylogenetic spread, 
a variety of genome sizes and average gene lengths as the model is increasingly unlikely to generalize 
well beyond the borders given by the training data. However, it also may be difficult 
for the model to learn and generalize if the genomic patterns inside the data are 
too different from each other. Practically, it was important for us to limit the size of
our training genomes set to be small enough that we could get experimental 
results within a couple of days and thereby be able to test many different data and 
model configurations.

We employed an iterative approach to effectively select a proper set of training 
genomes, starting from 3-4 genomes that were expected to be of the 
highest quality. We then evaluated the performance of the resulting model on all 
genomes individually (including the training genomes) and looked for candidates to 
add to the training set or to remove from it. This process was repeated multiple 
times concurrently with the model search until we saw no more room for substantial 
improvement given our computational constraints.

The genomes that were ultimately not used during training from the original 
split are hereafter refered to as the {\it validation genomes}. 
We report only the generalization 
performance on the combined set of validation and test genomes as the validation 
genomes are by far the largest set, and there was not a noticeable difference in 
performance between those two groups.

@@

\subsection{Data Pre-processing}
We first pre-processed and stored the raw genomic information by using GeenuFF 
(https://github.com/weberlab-hhu/GeenuFF). 
GeenuFF is a tool for checking and exploring genomic data and annotations, that 
stores all information inside a SQL database. Training and evaluation-ready data was
generated by querying this database and then transforming the returned data into a 
numerical format suitable for machine learning. The encoding of the genomic sequence 
was done in line with the IUPAC nucleic acid notation and the functional annotation 
used as labels during training is transformed to a one hot encoding with the four 
classes {\it Intergenic}, {\it UTR}, {\it CDS} and {\it Intron}. See Table 
\ref{tab:arrays} for a more detailed description of the generated data types.

@@

During data generation, we queried for the transcript with the longest protein of each gene and 
disregard FASTA sequences that have no functional annotation as it is ambiguous
whether such sequences contain no genes or 
were simply not annotated in the reference. GeenuFF checks the 
genomic annotations for potential errors during import and is able to mark those 
areas. We used this information to effectively mask those bases during training 
by using the sample weights described in Table \ref{tab:arrays}. The vast majority 
of masked bases lie in the intergenic region as the most prevalent error is a 
missing UTR and GeenuFF marks a potentially large intergenic region for it. Table 
\ref{tab:statistics_all} shows statistics about the masking.

For the training itself, we divided each continuous genomic sequence into 20,000 bp 
long subsequences, for which one-hot vectors of the base pairs and annotations 
are generated and respectively used as input and label for the neural network 
together with the sample weights. We appended zero padding if the subsequences  
are shorter than 20,000 bp. 
If a subsequence is fully marked as erroneous, we excluded it from all analyses.

@@

\subsection{Metrics}
\label{sec:metrics}
We mainly used two metrics to judge model performance. The {\it Genic F1} was selected as our 
primary metric and provides the most comprehensive picture of annotation quality 
in one number. The {\it Subgenic F1} is similar to the Genic F1, except that it does 
not take UTR predictions into account. This was calculated for comparability with AUGUSTUS 
as further explained in Section \ref{sec:augustus}. Both metrics work by summing up 
True Positives (TP), False Positives (FP) and False Negatives (FN) of each 
considered class before calculating precision and recall from these values and 
combining the two into a F1 score. Neither metric includes the
intergenic class, as this class is very abundant and appears to be by 
far the easiest to predict. (See Table \ref{tab:statistics_all} for the class 
distribution and Tables S\ref{suptab:generalization_animals} 
and S\ref{suptab:generalization_plants} for a more in-depth report our model performance 
including the intergenic class). In the case of the Genic F1, this means that 
effectively only the TP of the intergenic class were disregarded. The metrics 
essentially provide a weighted mean of the performances in the considered classes,
with weights proportional to class frequency.

@@

\subsection{Model Architecture and Training}
\label{sec:model}
We used a 4-layer deep stacked BLSTM network with 256 units per layer and layer 
normalization \citep{ba2016layer} between each BLSTM layer to produce the 
predictions in the form of a base pair wise classification. The model consists of 
circa 5.4 million parameters and was implemented with the deep learning library 
Keras (https://keras.io) on top of TensorFlow \citep{abadi2016tensorflow}. We 
tested multiple different architectures before arriving at this model configuration, 
including convolutional neural networks (CNN) and hybrid architectures. We also used 
class weights as the class frequencies are both unbalanced and 
vary greatly between animals and plants. 
For example, the trade-off between the number of intergenic and intronic bases is 
quite different in both groups (see Table \ref{tab:statistics_all}).

A class wise evaluation including the calculation of the Genic 
F1 score was performed on the validation set after every epoch. The best model
from any training run was selected as the model with the highest Genic F1 after either
maximum epochs were reached or Genic F1 stopped improving and training was interrupted.

@@

We also compared our final models to a dilated CNN (dCNN) 
baseline as this kind of architecture is 
widely used when working with DNA data as input. Details of the neural architecture 
search are given in Section \ref{ssec:dcnn} of the Supplements.

\subsection{Inference Techniques}
We also used multiple techniques to improve the prediction quality after the 
training was done. One very effective way for genomes with larger genes was inputting
longer sequences during inference than during training. This was done for all 
animal genomes except the invertebrates. The input sequences were up to 10 times 
longer, depending on the phylogenetic group and assembly quality. This also demonstrates the 
ability of our model to generalize as it is able to make successful prediction on 
far longer sequences than it has ever seen. The 
concrete lengths were chosen to keep the typical average gene length roughly 
proportional to the length of the sequence input. For more implementation details 
on this see Section \ref{ssec:longer} of the Supplements or the source code.

The final predictions of a single model were constructed by overlapping predictions, 
which were made from a sliding window and then cropped to a core sequence. This was done 
to strongly reduce a typical drop in performance of the models towards the beginning 
and end of each sequence (see Figure \ref{fig:main_results}c). It also improves the 
average model performance by providing the model with multiple different 
starting points. The different overlapping sequences were combined by averaging 
the individual softmax values of each base. 
The figures in Section \ref{ssec:overlapping} of the 
Supplements show the effect of overlapping for each genome, ordered by N75. We found 
that overlapping tends to work best if the genomes are not very fragmented and we 
used it for both animals and plants.

A model ensemble with 8 components was used to generate the final predictions for 
each species. For this, we performed 4 separate animal and plant training runs and 
selected two checkpoints each, specifically the checkpoint from the epoch with the best precision and the
other from the epoch with the best recall. The model checkpoint with the best Genic F1 was always best in one of 
those. This was done to increase the diversity of the model ensemble. As with the 
overlapping, the fusion of the 8 individual predictions was done by averaging the 
softmax values of each base pair prediction.

\subsection{Evaluation of AUGUSTUS} 
\label{sec:augustus}
Due to the practical time constraints for retraining and running, we decided to compare our Helixer
models only to one existing {\it de novo} tool, namely the popular gene caller AUGUSTUS \citep{stanke2003gene}.
Typical usage requires retraining AUGUSTUS for each species, with the exception of a
few lucky cases where a model for a sufficiently close relative is already available. To scale this for 
the large plant and animal datasets, we used protein-homology to create an AUGUSTUS training set and
therefore could train and evaluate only models without UTRs.

Orthologs of highly conserved generally single-copy genes were identified in each genome using
BUSCO. The {\it viridiplantae} set was used for plants and the {\it metazoa} set for animals.
The training genbank files were generated directly by BUSCO, by utilizing the ``--long"  % inline code?
parameter. For plants, the entire retraining could be performed as above; however, for animals
we randomly selected only half the BUSCO-generated training set, which resulted in a training set 
size and runtime comparable to the plants (about 400 genes and several CPU days per species). For animals
the training was carried out with the subsetted training file and using the following scripts provided
by AUGUSTUS. An untrained model was setup with ``new\_species.pl", and the model was fit by running
"etraining" before and after the major hyper-parameter optimization with ``optimize\_augustus.pl".
Using the trained model for each species, we ran the main prediction ("augustus") with ``--UTR=off"
and ``--gff3=on". The gffs produced by AUGUSTUS were imported into GeenuFF and exported as HDF5
files in the same manor as the reference, allowing for the direct comparison with both the reference and 
Helixer predictions. While this method was feasible for some 237 species, it allowed neither training
nor prediction of UTR regions with AUGUSTUS, so the metric Subgenic F1, which disregards the class UTR,
was used for all comparisons between AUGUSTUS and Helixer (see Section \ref{sec:metrics} for more details).

@@

For each of these six species the following search was performed on Sequence Read Archive \citep{leinonen2010sequence}
`(("<species name>"[Organism] OR <species name>[All Fields]) AND stranded[All Fields]) 
AND ("biomol rna"[Properties] AND "library layout paired"[Properties])'.
If more than 50 samples were identified, every Nth samples was selected 
so that in total under 50 samples were chosen for further processing
(see Table S\ref{suptab:rnaseq_samples}).
Each sample was prepped, mapped, and quality controlled in a pipeline using Trimmomatic 
\citep{bolger2014trimmomatic}, 
Hisat2 \citep{kim2019graph}, Samtools \citep{li2009sequence}, PicardTools 
(http://broadinstitute.github.io/picard/), 
FastQC 
(www.bioinformatics.babraham.ac.uk/projects/fastqc/),
and MultiQC (\cite{ewels2016multiqc}; see
Table S\ref{suptab:bioinfo_params} for details). This pipeline was 
automated and the results visualized with the code available here (https://github.com/weberlab-hhu/RNAsleek).
The samples were filtered to those with relatively high mapping rates, high mapping to
exonic relative to non-exonic regions, low 3' bias, normal FastQC and Trimmomatic stats, 
and a stranded mapping pattern (2nd read is sense strand). If more remained, seven
of the high quality samples were selected randomly.

@@

\begin{figure*}[!tpb]
\label{fig:main_results}
\centerline{\includegraphics[width=\textwidth]{images/main_results}}
\caption{Main overall results and other investigations. The top and bottom rows
show the animal and plant data, respectively. Orange dots are for AUGUSTUS and 
the blue ones are for Helixer. {\bf a)} Swarm - and boxplots showing the Subgenic F1 
scores for all validation and test genomes. Each dot represents the prediction 
performace for one genome. The median prediction scores are shown by the red line. 
Details of how AUGUSTUS was used are given in Section \ref{sec:augustus}. {\bf b)} 
Scatterplots showing the prediction performances measured in Subgenic F1 by genome 
length. The regression line is shown with a 95\% confidence interval. {\bf c)} The 
average Genic F1 (in red) and basepair wise accuracy (in purple) with respect to 
the position in the 20,000 basepair long input sequences. The dashed line shows the 
same without overlapping. Each value is the average performance of a 200 basepair 
long subsection by one of the models of the final ensemble. 
See Section \ref{ssec:overlapping} of the 
Supplements for the effects of overlapping in individual species.}
\end{figure*}

@@

\section{Results}
Figure \ref{fig:main_results}a shows the side-by-side comparison of the distribution 
of performances on all non-training genomes in the animal and plant case by Subgenic 
F1. In the case of Helixer, these scores represent cross-species predictions. We 
also compare the median performances of different configurations of Helixer with 
AUGUSTUS and a dilated CNN baseline in Table \ref{tab:results}. The results show a 
clear improvement over the AUGUSTUS both in higher median performance and reduced spread. We also 
significantly outperform a baseline constructed of a dilated CNN architecture.

Our models, however, tend to perform less well for the very smallest and largest 
genomes or species that are phylogenetically the furthest away from our training 
genomes. This is the case for both animals and plants and is visualized in
Figures S\ref{supfig:phylogeny_f1_animals} and S\ref{supfig:phylogeny_f1_plants}. 
We do not, for example, consistently predict very well on the 
algae nor on the non-avian reptiles. While one of the algae were included 
in our training genomes it accounted for a tiny proportion of the total training 
data and the non-avian reptiles were not included at all.

@@

Two techniques were used during inference to improve performance and limit model 
bias. The usage of longer input lengths helped especially for genomes that tend to 
have longer genes and was enabled by our model architecture being a relatively 
simple BLSTM stack without any fully connected layers on top. We also constructed the 
final predictions out of a overlapping ones, which greatly helped to reduce 
prediction bias in most genomes. To our knowledge, neither technique has been used 
before in a model developed for gene annotation.

@@

To make an independent qualitative and quantitative evaluation of
performance we compared Helixer predictions, AUGUSTUS predictions and the
reference to RNAseq coverage data. Evaluation with RNAseq data was performed for
three each of animal and plant species. RNAseq coverage provides support for an 
exonic annotation (UTR or CDS), spliced coverage provides support for an 
annotation of intron, and neither coverage nor spliced coverage is expected 
for intergenic annotations. Looking at selected subsequences (Figure \ref{fig:cov_example_main}) % and supplemental figures.
we identified cases where RNAseq supported a) both the annotations of the reference 
and Helixer, b) neither, c) the reference, but not Helixer,
and d) Helixer, but not the reference.

@@

Helixer models also sometimes showed uncertainty for larger regions.
In some cases where Helixer did not receive RNAseq support for its highest probability
annotation, it assigned a low but non-trivial probability to the RNAseq-supported 
exon/intron pattern (Figures \ref{fig:cov_example_main}b-c, S\ref{supfig:cov_example_02}c). However, in the extreme,
there are cases where the Helixer model exhibits substantial indecision or confusion 
and shifts gradually between classes with no single class receiving a high probability 
for extended stretches (Figures S\ref{supfig:cov_example_02}c-d). 
Notably, in one of the examples (Figure S\ref{supfig:cov_example_02}d)
the RNAseq shows evidence of alternative splicing; 
and in another (Figure S\ref{supfig:cov_example_02}c) Helixer's
prediction falls between that of the reference and of AUGUSTUS.
